# Functions

## Reading on functions from RDS by Hadley Wickham

[Functions from RDS](https://r4ds.hadley.nz/functions){target="_blank"}

## Tutorial 

This is a good opportunity to illustrate the usefulness of writing your own functions. When you install packages in R, you get a bunch of functions you can use. But you can also create these on your own to simplify your analyses!

You do this with the following syntax: MyNewFunction <- function(param1, param2){ >code >}

Whatever the last line of the “code” portion of the function spits out, get’s returned from the function. So if you said X <- mynewfunction(param1, parm2) X would now have it in whatever your function returned. See a simple example below: a function that adds 1 to any number we pass to it.


```{r}

add1 <- function(number){
          number + 1
          }

add1(14)
```

You can make a generic ggplot function. Here I make a geom_point function. You can adjust the ggplot details to whatever you want. 


```{r}

library(tidyverse)

```


```{r}

point_fun <- function(...) {
  ggplot(df, aes(x = x, y = y, color = color)) +
    geom_point() +
    labs(x = x_lab, y = y_lab, title = title, color = color_lab)
}


```

Then you just tell it what the inputs are. If you don't want to color by something you need to remove that from the function. If you don't want a title you just say title <- ""

Of course you could make a variety of these. A line function. So on. 

The mtcars is just an example df that comes with R for demonstration. 
```{r}
df <- mtcars
x <- mtcars$hp
y <- mtcars$mpg
x_lab <- "Horsepower"
y_lab <- "MPG"
color <- mtcars$cyl
color_lab <- "Cyl"
title <- "Plot title"
  
point_fun()

```

You could also save the plot as an object and adjust. 
```{r}

p <- point_fun()

p + theme_bw() + geom_line() + scale_x_reverse()

```

In general you should write a function for anything that you do repetitively. So if you have a particular analysis that you will do many times, rather that cutting and pasting or writing the code over and over again, just write a function, save that function, and call it when you need it. 

Next week will we look at for loops which are also useful for iterating something over and over again. 


