# For loops


Here we are going to learn how to write a for loop to download some data and name files. We will do two sites for simplicity, but this can be fed as many sites as you wish!

## Libraries
```{r}
library(tidyverse)
library(dataRetrieval)

```

## Defining parameters 

You could type out the sites and names or you could pull it from some whatNWISsites call. 
```{r}
sites <- c("06037500", "12363000")
names <- c("salmon", "flathead")
params <- c("00060", "00010") # This is temperature and Q. You can obviously change and then also change the readNWISdata to something else and the parameters, etc. 
start <- as_date("1900-10-01") # Just a nice round number for start time. 
end <- as_date("2022-09-30") #End of last water year. 
service <- "dv" # Daily values

# Removed those sites for testing
# "12363000", "13296000", "12344000", "12447200", "13331500", "06191500"

```

## Writing the for loop 

Instead of writing a function, we use for(i in sites). So this will run through the list of sites and download the data and do whatever else you tell it. Then we create a filename that is data_, i. So it will write out data_06037500.csv for the first, data_12363000.csv for the second, so on. Then I just read the .csv in to check and make sure it worked.

```{r}
for (i in sites) {
  tmp <- readNWISdata(sites = i, parameterCd = params, service = service, startDate = start, endDate = end) %>% 
    renameNWISColumns()
  filename <- paste("data_", i, sep="")
  write_csv(tmp, file = paste(filename, ".csv", sep=""))
}

test_df <- read_csv("data_06037500.csv")
```


## Adding second for

Here, I've added a second for. For i in sites and for j in names. So when we get to the naming we look to the j for the name to put that in the title of the .csv
```{r}
for (i in sites) {
  for (j in names) {
  tmp <- readNWISdata(sites = i, parameterCd = params, service = service, startDate = start, endDate = end) %>% 
    renameNWISColumns()
  filename <- paste("data_", j, sep="")
  write_csv(tmp, file = paste(filename, ".csv", sep=""))
  }
}

test_df <- read_csv("data_salmon.csv")
```

## Removing the tmp file

This does the same thing, just removes the tmp file at the end to clean up the global environment.

```{r}
for (i in sites) {
  for (j in names) {
  tmp <- readNWISdata(sites = i, parameterCd = params, service = service, startDate = start, endDate = end) %>% 
    renameNWISColumns()
  assign(paste("data_", j, sep=""), tmp)
  rm(tmp) 
  }
}
```
