# daymetr

The daymetr package can be used to download daymet climate data. 

[See here](https://cran.r-project.org/web/packages/daymetr/vignettes/daymetr-vignette.html){target="_blank"}

[And here](https://tmieno2.github.io/R-as-GIS-for-Economists/daymet-with-daymetr-and-feddata.html){target="_blank"}

For some daymetr tutorials. 


## Install Packages

```{r}
# Install and load required packages
# install.packages("daymetr") #uncomment here to install the package. 
library(daymetr)
library(tidyverse)

```


## Download daymet for a specific location. 
```{r}
df <- download_daymet(
  site = "Oak Ridge National Laboratories",
  lat = 36.0133,
  lon = -84.2625,
  start = 2000,
  end = 2010,
  internal = TRUE,
  simplify = TRUE # return tidy data !!
  ) 
```

## Here are the daymet variables:

..$ year         : int [1:1095] 2000 2000 2000 2000 2000 2000 2000 2000 2000 2000 ...
  ..$ yday         : int [1:1095] 1 2 3 4 5 6 7 8 9 10 ...
  ..$ dayl..s.     : num [1:1095] 34571 34606 34644 34685 34729 ...
  ..$ prcp..mm.day.: num [1:1095] 0 0 0 0 0 0 0 0 0 0 ...
  ..$ srad..W.m.2. : num [1:1095] 334 231 148 319 338 ...
  ..$ swe..kg.m.2. : num [1:1095] 18.2 15 13.6 13.6 13.6 ...
  ..$ tmax..deg.c. : num [1:1095] 20.9 13.88 4.49 9.18 13.37 ...
  ..$ tmin..deg.c. : num [1:1095] -2.73 1.69 -2.6 -10.16 -8.97 ...
  ..$ vp..Pa.      : num [1:1095] 500 690 504 282 310 ...
 - attr(*, "class")= chr "daymetr"
 
Those are day length in seconds, precip in mm/day, solar radiation in w/m2, swe in kg/m2, tmax in C, tmin in C, and vapor pressure in Pascals. 
 
## Make a date column and plot max temperature.  
```{r}

df <- df %>% 
  mutate(date = as.Date(paste(year, yday, sep = "-"), "%Y-%j"))

df %>%
  filter(measurement == "tmax..deg.c.") %>%
  ggplot(aes(x = date, y = value)) +
  geom_line(color = "red") +
  theme_linedraw() +
  labs(y = "Max T (C)", x = "")
```
 
## Do the same for radiation.  
```{r}
df %>%
  filter(measurement == "srad..W.m.2.") %>%
  ggplot(aes(x = date, y = value)) +
  geom_line(color = "red") +
  theme_linedraw() +
  labs(y = "Solar radiation (W/m2)", x = "")
```

## Stack 2 variables with patchwork. 

[Here is a refernce for special characters in ggplot axes labels](https://www.casualinferences.com/posts/adding-special-characters-ggplot-labels/){target="_blank"}

```{r}
library(patchwork)

temp_plot <- df %>%
  filter(measurement == "tmax..deg.c.") %>%
  ggplot(aes(x = date, y = value)) +
  geom_line(color = "red") +
  theme_linedraw() +
  labs(y = "Max T (\u00b0C)", x = "")

srad_plot <- df %>%
  filter(measurement == "srad..W.m.2.") %>%
  ggplot(aes(x = date, y = value)) +
  geom_line(color = "red") +
  theme_linedraw() +
  labs(y = "Solar radiation"~(W/m^2), x = "")

srad_plot / temp_plot

```






